Coherence of Molecular Mechanisms In Major Human Disease and Traits

### 2020-04-05 manuscript/submission.BMC_Bioinformatics/Revision1/manuscript_BMC_Bioinformatics_R1.Rmd
Tweak title
Latest changes for R1 submission

---
bibliography: [references/networks.bib,references/brain.bib,references/enrichment.bib,references/3D_refs.bib]
csl: styles.ref/bmc-bioinformatics.csl
output:
  word_document:
    reference_docx: styles.doc/BMC_Bioinformatics.docx
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Coherence of Molecular Mechanisms <!--in Genes Identified Through GWAS -->In Major Human Disease and Traits

@@

**Background.** Complex phenotypes such as height and intelligence, are thought to be a product of the collective effects of multiple phenotype-associated genes, or, more precisely, interactions among their protein products. High/low degree of interactions is suggestive of coherent/random molecular mechanisms, respectively. Comparing the degree of interactions may help to better understand the coherence of molecular mechanisms underlying complex phenotypes and the potential for therapeutic intervention. However, direct comparison of the degree of interactions is difficult due to different sizes and configurations of phenotype-associated gene networks.

**Methods.** We introduce a measure of network coherence as a slope of internal vs. external distributions of the degree of interactions. The internal degree distribution is defined by interaction counts within a phenotype-specific gene network, while the external degree distribution counts interactions with other genes in the whole protein-protein interaction (PPI) network. We present a novel method for normalizing the coherence estimates, making them directly comparable.

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 116 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the reference for high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric disorders and intelligence-related traits were the least coherent.

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of phenotype-associated gene networks that accounts for the network size and shape differences. Our results highlight gaps in our current knowledge of genetics and molecular mechanisms of complex phenotypes and suggest priorities for future GWASs.

@@

Genome-wide association studies (GWAS) have significantly advanced our understanding of complex phenotypes by identifying disease- and trait-associated genetic markers and suggesting corresponding genes [@Altshuler:2008aa; @Risch:1996aa]. However, GWAS findings explain only a fraction of heritability [@Eichler:2010aa; @Maher:2008aa; @Manolio:2009aa]. Furthermore, the corresponding genes are often spread across different chromosomes with no known connection to one another, hindering understanding of the molecular mechanisms. These limitations of current generation GWAS might stem from the yet incomplete knowledge of genetic determinants of complex phenotypes, the potential heterogeneity of complex phenotypes and/or the causes other than genetics [@Botstein:2003aa; @turkheimer2012genome; @Sullivan:2018aa].

Many studies have shown that genetically-driven complex phenotypes are often associated with functionally related genes that are more likely to form networks of interacting protein products [@Hartwell:1999aa; @Krauthammer:2004aa; @Huang:2006aa; @Lage:2007aa; @Cordell:2009aa; @Emily:2009aa; @Feldman:2008aa]. This observation has been verified systematically for a large number of diseases [@Menche:2015aa], thus confirming a fundamental hypothesis of the interactome-based approach to understanding human phenotypes, namely that disease- and trait-associated genes tend to form interaction modules [@Goh:2007aa]. Such interaction modules may be viewed as connected subnetworks within the full interactome and may contain all molecular determinants of a certain phenotype. Consequently, methods to identify phenotype-associated networks of functionally-related genes have been developed [@Krauthammer:2004aa; @Cordell:2009aa; @Akula:2011aa].

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by interactions among phenotype-associated genes, or, more precisely, their protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) that form an interaction network are more likely to share similar functions [@Farber:2013aa], or co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes forming networks of different sizes and configurations.

<!--This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the slope of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absence of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.-->
This study presents a novel measure of network coherence as a slope of the internal versus external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used selected gene sets from the MSigDb database [@Subramanian:2005aa] as a reference for highly coherent networks, while sets of randomly sampled genes were used as the reference for the absence of coherence. Using MSigDb and random networks matched in size to phenotype-specific networks, we derived a normalized measure of coherence that can be compared across phenotypes. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of complex phenotypes, such as intelligence-related traits, to have low coherence and metabolic traits and diseases to have high coherence. Our method enables direct comparison of coherence measures and highlights gaps in the current understanding of molecular mechanisms of many phenotypes, e.g., Major Depressive Disorder having the lowest coherence in the already low-coherence "Psychiatric disease" category.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensures that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs were removed from the analysis (see Methods) leaving 116 phenotypes that could be analyzed. Second, full STRING data was used to maximize the use of PPI information at the expense of potential noise (referred to hereafter as “STRING”). The advantage of using complete data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs having low confidence scores. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, was used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed substantial negative correlation between network sizes and the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Additional File 2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

This work was supported by a Genetics and Human Agency Award from the Templeton Foundation to KSK, MR, CC, SB. The content is solely the responsibility of the authors and does not necessarily represent the official views of the Templeton Foundation.

@@

To alleviate the dependency of degree distributions from the size of phenotype-associated gene sets and the content of a selected PPI database, we normalized slopes of phenotype-specific networks to the range of slopes formed by degree distributions of highly coherent and random networks. Specifically, we select 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections, $\beta_{min}$ median slope) and 1,000 random networks ($\beta_{max}$ median slope) matched in size to the corresponding network of phenotype-specific genes. For each phenotype $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of MSigDb networks. These values represent normalized coherence and can be compared across phenotypes and PPI databases.

@@

**Additional File 3. Phenotype-specific coherence estimates.** The PPI database (STRING, STRING filtered, Biogrid) are specified in the corresponding column names. For each phenotype category, the results are sorted by the "String Filt Coherence" column in descending order. "NA" indicates the corresponding value cannot be estimated due to lack of sufficient number of genes annotated with PPIs.

**Additional File 4. Normalization of coherence estimates alleviates its network size dependence.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the number of SNPs, genes, and the normalized coherence estimates using the corresponding PPI databases.

>>>>

---
bibliography: /Users/mdozmorov/Documents/Work/presentations/Templates/manuscript_template/references/references_MD.bib
csl: ../../styles.ref/bmc-bioinformatics.csl
output:
  word_document:
    reference_docx: ../../styles.doc/BMC_Bioinformatics.docx
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

# <!-- New in Revision 1 -->A Method for Estimating <!-- End New in Revision 1 -->Coherence of Molecular Mechanisms <!--in Genes Identified Through GWAS -->In Major Human Disease and Traits

@@

**Background.** Phenotypes such as height and intelligence, are thought to be a product of the collective effects of multiple phenotype-associated genes and interactions among their protein products. High/low degree of interactions is suggestive of coherent/random molecular mechanisms, respectively. Comparing the degree of interactions may help to better understand the coherence of phenotype-specific molecular mechanisms and the potential for therapeutic intervention. However, direct comparison of the degree of interactions is difficult due to different sizes and configurations of phenotype-associated gene networks.

**Methods.** <!-- New in Revision 1 -->We introduce a metric for measuring coherence of molecular-interaction networks<!-- End New in Revision 1 --> as a slope of internal vs. external distributions of the degree of interactions. The internal degree distribution is defined by interaction counts within a phenotype-specific gene network, while the external degree distribution counts interactions with other genes in the whole protein-protein interaction (PPI) network. We present a novel method for normalizing the coherence estimates, making them directly comparable.

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 116 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the reference for high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric disorders and intelligence-related traits were the least coherent. <!-- New in Revision 1 -->We demonstrate that coherence and modularity measures capture distinct network properties.<!-- End New in Revision 1 -->

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of molecular-interaction gene networks that accounts for the network size and shape differences. Our results highlight gaps in our current knowledge of genetics and molecular mechanisms of complex phenotypes and suggest priorities for future GWASs.

@@

Genome-wide association studies (GWAS) have significantly advanced our understanding of complex phenotypes by identifying disease- and trait-associated genetic markers and suggesting corresponding genes [@Altshuler:2008aa; @Risch:1996aa]. However, GWAS findings explain only a fraction of heritability [@Eichler:2010aa; @Maher:2008aa; @Manolio:2009aa]. Furthermore, the corresponding genes are often spread across different chromosomes with no known connection to one another, hindering understanding of the molecular mechanisms. These limitations of current generation GWAS might stem from the yet incomplete knowledge of genetic determinants of complex phenotypes, the potential heterogeneity and/or the causes other than genetics [@Botstein:2003aa; @turkheimer2012genome; @Sullivan:2018aa].

Many studies have shown that genetically-driven phenotypes are often associated with functionally related genes that are more likely to form networks of interacting protein products [@Hartwell:1999aa; @Krauthammer:2004aa; @Huang:2006aa; @Lage:2007aa; @Cordell:2009aa; @Emily:2009aa; @Feldman:2008aa]. This observation has been verified systematically for a large number of diseases [@Menche:2015aa], thus confirming a fundamental hypothesis of the interactome-based approach to understanding human phenotypes, namely that disease- and trait-associated genes tend to form interaction modules [@Goh:2007aa]. Such interaction modules may be viewed as connected subnetworks within the full interactome and may contain all the molecular determinants of a certain phenotype. Consequently, methods to quantitatively assess these phenotype-associated networks of functionally-related genes have been developed [@Krauthammer:2004aa; @Cordell:2009aa; @Akula:2011aa].

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by interactions among phenotype-associated genes, or, more precisely, their protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) forming an interaction network are more likely to have similar functions [@Farber:2013aa] and co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes forming networks of different sizes and configurations.

<!--This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the slope of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absence of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.-->
This study presents a novel measure of network coherence as a slope of the internal versus external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used selected gene sets from the MSigDb database [@Subramanian:2005aa] as a reference for highly coherent networks, while sets of randomly sampled genes were used as the reference for the absence of coherence. Using MSigDb and random networks matched in size to phenotype-specific networks, we derived a normalized measure of coherence that can be compared across phenotypes. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of intelligence-related traits to have low coherence and metabolic traits and diseases to have high coherence. Our method enables direct comparison of coherence measures and highlights gaps in the current understanding of molecular mechanisms of many phenotypes, e.g., Major Depressive Disorder having the lowest coherence in the already low-coherence "Psychiatric disease" category.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We used three types of PPI data. First, STRING data filtered by interaction confidence score above 500 (the middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step selects for only high-confidence PPIs. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that, for some phenotypes, the number of PPIs may be insufficient for forming networks. Consequently, these phenotypes were removed from the analysis (see Methods) leaving 116 phenotype-associated gene sets that could be analyzed. Second, full STRING data (henceforth, "STRING") was used to maximize the use of PPI information at the expense of potential noise. The advantage of using complete data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs with low confidence scores. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, was used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed substantial negative correlation between network sizes and the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Additional File 2). <!-- New in Revision 1 -->Similarly, slopes of the phenotype-specific networks and sizes of the corresponding confidence intervals negatively correlated with the number of genes in the networks (Average PCC -0.49/-0.70, respectively, Additional File 3, Additional File 4). This is expected as network size affects internal degree distributions and the corresponding slopes on the internal vs. external degree plots, with the larger networks allowing for a more precise estimation of slopes (smaller confidence intervals). <!-- End New in Revision 1 -->These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

This work was supported by a Genetics and Human Agency Award GF13361-152627 from the Templeton Foundation to KSK, MR, CC, SB. The content is solely the responsibility of the authors and does not necessarily represent the official views of the Templeton Foundation.

@@

To alleviate the dependency of degree distributions from the size of phenotype-associated gene sets and the content of a selected PPI database, we normalized slopes of phenotype-specific networks to the range of slopes formed by degree distributions of highly coherent and random networks. Specifically, we select 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections, $\beta_{min}$ median slope) and 1,000 random networks ($\beta_{max}$ median slope) matched in size to the corresponding network of phenotype-specific genes. For each phenotype $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of MSigDb networks. These values represent normalized coherence and can be compared across phenotypes and PPI databases. <!-- New in Revision 1 -->In addition, for each coefficient $\beta_i$, we extract confidence intervals and normalize them as the coefficients themselves. Note the normalization step alters common interpretation of confidence intervals; therefore, results of the permutation test of slopes should be used to compare coherences.<!-- End New in Revision 1 -->

@@

**Additional File 3. Phenotype-specific coherence estimates.** The PPI database (STRING, STRING filtered, Biogrid) are specified in the corresponding column names. For each phenotype category, the results are sorted alphabetically. <!-- New in Revision 1 -->In addition to the normalized coherence and the permutation p-values indicating significant difference of phenotype-specific coherence from random, modularity, untransformed slopes, and the corresponding lower and upper confidence interval bounds are shown. <!-- New in Revision 1 -->"NA" indicates the corresponding value cannot be estimated due to lack of sufficient number of genes annotated with PPIs.

**Additional File 4. Normalization of coherence estimates alleviates its network size dependence.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the number of SNPs, genes, normalized coherence estimates, <!-- New in Revision 1 -->and the untransformed slopes and the sizes of their confidence intervals (CI)<!-- End New in Revision 1 --> using the corresponding PPI databases.
